File-based localization of numerical perturbations in data analysis pipelines

### 2020-02-16 paper.tex
Updates

Operating systems and other computational factors are known to have an
effect on the results produced by neuroimaging pipelines, presumably due to
the creation, propagation and amplification of numerical errors along data
analysis pipelines. However, the precise causes of such instabilities and
the path along which they propagate in the pipelines are unclear, while
this process could play a major role in the current reproducibility crisis.
We present a technique to identify the computational processes in a
pipeline that create numerical differences when executed in different
computational conditions. Our technique leverages system-call interception to reconstruct
processing and file graphs without pipeline instrumentation. Applying our tool to
the structural pre-processing pipelines released by the Human Connectome
Project, we find that linear and non-linear registration are the cause of
most numerical instabilities in such pipelines, which confirms previous
findings.

@@

In neuroimaging, our primary application field, data analyses often consist
of hundreds of computational processes, sometimes coming from multiple
toolboxes, that are aggregated to perform a specific function. For
instance, the fMRIprep pipeline~\cite{esteban2019fmriprep} assembles software blocks
from FSL, AFNI, Freesurfer and ANTs to provide a state-of-the art
functional MRI processing tool with minimal user input. Another example are the pipelines of the Human
Connectome Project~\cite{glasser2013} that combine tools from FSL and Freesurfer to pre-process
structural, functional and diffusion data from their
open dataset. In both cases, the pipelines leverage toolboxes that are
widely trusted in the community. At the same time, substantial results
variations have been observed in these toolboxes resulting from minor data
or infrastructure perturbations~\cite{Glatard2015, Gronenschild2012, Lewis2017-ll}, which suggests that further investigation of their
numerical conditioning is required. For such complex pipelines, a
lightweight solution has to be found to perform such evaluations with
limited code instrumentation.

>>>>

Data analysis pipelines are known to be impacted by operating system
updates and other infrastructural factors, presumably due to the creation,
propagation and amplification of numerical errors. However, the precise
causes of such instabilities and the path along which they propagate in the
pipelines are unclear, while this process could play a major role in the
current reproducibility crisis. We present a technique to identify the
computational processes in a pipeline that create numerical differences
when executed in different computational conditions. Our technique
leverages system-call interception to reconstruct processing and file
graphs without pipeline instrumentation. By applying our tool to the
structural pre-processing pipelines released by the Human Connectome
Project, we find that linear and non-linear registration are the cause of
most numerical instabilities in such pipelines, which confirms previous
findings.

@@

In neuroimaging, our primary application field, data analyses often consist
of hundreds of computational processes, sometimes coming from multiple
toolboxes, that are aggregated to perform a specific function. For
instance, the fMRIprep pipeline~\cite{esteban2019fmriprep} assembles software blocks
from FSL~\cite{jenkinson2012fsl}, AFNI~\cite{cox2012afni}, Freesurfer~\cite{fischl2012freesurfer} and ANTs~\cite{avants2009advanced} to provide a state-of-the art
functional MRI processing tool with minimal user input. Another example are the pipelines of the Human
Connectome Project~\cite{glasser2013} that combine tools from FSL and Freesurfer to pre-process
structural, functional and diffusion data from their
open dataset. In both cases, the pipelines leverage toolboxes that are
widely trusted in the community. At the same time, substantial results
variations have been observed in these toolboxes resulting from minor data
or infrastructure perturbations~\cite{Glatard2015, Gronenschild2012, Lewis2017-ll}, which suggests that further investigation of their
numerical conditioning is required. For such complex pipelines, a
lightweight solution has to be found to perform such evaluations with
limited code instrumentation.
